|

### 2021-07-24 prova_knit_finale.Rmd
qualche modifica

\section{State-Space Models}
We now define \textit{State-Space models} as Markov processes that are partially observed. Let $X_t\in\X$ and $Y_t\in\Y$ and let $\{(X_t,Y_t)\}$ be a stochastic process. To help intuition, suppose that $\{Y_t\}$ is observable, i.e., coincides with a sequence of observations, while $\{X_t\}$ is latent. Consider the initial distribution $\P_0(dx_0)$ and the probability kernels
$P_t(x_{t-1},dx_t)$ and $F_t(x_t,d_t)$, where $t=1,2,\dots$.
\
\begin{definition}\label{state-space-def}
The process $\{(X_t,Y_t)\}$ is a State-Space model if the joint distribution of $(X_{0:T},Y_{0:T})$ is 
\begin{align*}
  \P_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T}) &= \P_0(dx_0)\prod\limits_{t=1}^T P_t(x_{t-1},dx_t)\prod\limits_{t=1}^T F_t(x_t,dy_t)\\
  &= [\P_0(dx_0)F_0(x_0,y_0)]\prod\limits_{t=1}^T[P_t(x_{t-1},dx_t)F_t(x_t,dy_t)] \\
  &= \P_T(dx_{0:T})\prod \limits_{t=1}^T F_t(x_t,dy_t).
\end{align*}
\end{definition}
From the second inequality, it is apparent that $\{(X_t,Y_t)\}$ is a Markov process with initial distribution $[\P_0(dx_0)F_0(x_0,y_0)]$ and transition kernel $P_t(x_{t-1},dx_t)F_t(x_t,dy_t).$ The third equality, instead, shows that, $\{X_t\}$ has a marginal distribution $$\P_T(dx_{0:T})=\P_0(dx_0)\prod_{t=1}^T P_t(x_{t-1},dx_t),$$ meaning that $\{X_t\}$ is marginally distributed as a Markov process with initial distribution $\P_T(dx_{0:T})$ and transition kernel $P_t(x_{t-1},dx_t).$

Assume now that there exists a dominating measures $\nu(dy)$ such that 
$F_t(x_t,dy_t)=f_t(y_t\mid x_t)\nu(dy_t)$ for all $t.$
Then by the last equality of definition \ref{state-space-def}, the joint distribution of $(X_{0:T},Y_{0:T})$ is 
\begin{equation}\label{state-space-densities}
   \P_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T})=\P_T(dx_{0:T})\prod\limits_{t=0}^T f_t(y_t\mid x_t)\prod\limits_{t=0}^T\nu(dy_t).
\end{equation}
Integrating away $X_{0:T},$ one gets the marginal density
\begin{align*}\label{state-space-marginal-y}
  \P_T(dy_{0:T}) &= \E_{\P_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\nu(dy_s) \\
                &= p_t(y_{0:t})\prod\limits_{s=0}^t\nu(dy_s),
\end{align*}                
where $p_t(y_{0:t})$ is the likelihood function.
Dividing the left-hand side of equation \ref{state-space-densities} by the left-hand side of equation \ref{state-space-marginal-y} and the right-hand side of equation \ref{state-space-densities} by the right-hand side of equation \ref{state-space-marginal-y}, one obtains the filtering distribution
\begin{equation}
\P_t(X_{0:t}\in dx_{0:t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\P_t(dx_{0:t}).
\end{equation}

@@

Suppose that $\X=\R^{d_x}$ and $\Y=\R^{d_y}.$ Let $U_t\thicksim\N(0,\Sigma_t)$ and $V_t\thicksim\N(0,R_t)$ be two independent white-noise errors and assume the initial distribution 
$X_0\thicksim\N(0,\Sigma_0).$ We have a linear Gaussian model if 
\begin{align}
X_t &= A_tX_{t-1}+U_t \\
Y_t &= B_tX_t + V_t.
\end{align}
Note that  $X_t\mid X_{t-1}$ and $Y_t\mid X_t$ are both normally distributed. 
\
\begin{lemma}\label{kalman-filter}
  Let $X\thicksim\N(m_0,C_0)$ and $Y\mid X\thicksim\N(BX,R)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
  Then, it holds that
  \begin{align}
    X\mid Y=y&\thicksim\N(m_1,C_1),\\
    C_1&= C_0 (I_{d_x} - B'(B C_0 B'+ R)^{-1} B C_0) \\
    m_1&=[I_{d_x} - C_0 B'(B C_0 B'+ R)^{-1} B ]m_0 + C_0 B'(B C_0 B'+ R)^{-1}y.
  \end{align}
\end{lemma}
\begin{proof*}
   The lemma can be proved using Bayes' Theorem. Suppose $X\thicksim\N(m_0, C_0).$ It is easy to see that
\begin{equation} \label{normal-distrib-rewrite}
        p(x)\propto \text{exp}\Big(-\frac{1}{2}X'C_0^{-1}X + X'C_0^{-1}m_0\Big).
\end{equation}
Now, by Bayes' Theorem,

    \begin{align*}
        p(x\mid Y=y)  &\propto p(x) p(y\mid x) \\
                    &\propto \text{exp}\Big(-\frac{1}{2}X' C_0^{-1}X + X' C_0^{-1}m_0-\frac{1}{2}(y-BX)'R^{-1}(y-BX)) \\
                    &\propto \text{exp}(-\frac{1}{2}X' C_0^{-1}X + X' C_0^{-1}m_0-\frac{1}{2}X' B' R^{-1}BX+X' B' R^{-1}y)\\
                    &=\text{exp}(-\frac{1}{2}X'(C_0^{-1}+B' R^{-1}B)X+
                    X'(C_0^{-1}m_0+B' R^{-1}y)\Big).
    \end{align*}

Hence, it follows that 
\begin{equation*}
    C_1=Var(X\mid Y=y)= (C_0^{-1}+ B' R^{-1} B)^{-1}=C_0-C_0 B'(B C_0 B'+R)^{-1}B C_0
\end{equation*}
and 
    \begin{align*}
        m_1 &=\E(X\mid Y=y) = C_1^{-1}(C_0^{-1}m_0+B' R^{-1}y)\\
            &= (C_0-C_0B'(B C_0B'+R)^{-1}B C_0)(C_0^{-1}m_0+B' R^{-1}y)\\
            &= m_0-C_0B'(B C_0B'+R)^{-1}B m_0 + C_0 B' R^{-1}y - C_0B'(B C_0B'+R)^{-1}B C_0 B' R^{-1}y \\
            &=[I_{d_x} - C_0 B'(B C_0 B'+ R)^{-1} B ]m_0 + C_0 B'(B C_0 B'+ R)^{-1}y.
    \end{align*}
\end{proof*}

Assume that $\P_{t-1}(dx_{t-1}\mid Y_{1:t-1}=y_{1:t-1})=\N(m_{t-1},C_{t-1}).$
Then, using the transition equation of the latent variable, 
by the linearity of the Gaussian distribution, we have that the predictive distribution is 
\begin{equation}
\P_{t-1}(dx_{t}\mid Y_{1:t-1}=y_{1:t-1})=\N(A_tm_{t-1},A_tC_{t-1}A_t'+\Sigma_t).
\end{equation}
Letting $E_T=A_tC_{t-1}A_t'+\Sigma_t,$ and applying lemma \ref{kalman-filter}, it follows that the filtering distribution is 
\begin{align*}
  \P_{t}(X_t\in dx_{t}\mid Y_{1:t-1}&=y_{1:t-1})=\N(m_t,C_t), \text{where} \\ C_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
  m_t &= [I_{d_x} - E_tB_t'(B_tE_tB_t'+R_t)^{-1} B_t]A_tm_{t-1}+E_tB_t'(B_tE_tB_t'+R_t)^{-1}y_t.
\end{align*}

In some situations, it is possible to linearize otherwise non-linear models in order to employ this filtering method.
There are, however, cases where this is not possible or where the linearization is not satisfactory. One such example is the stochastic volatility model. In the following chapters, we will introduce powerful algorithms that can be used in much more general settings.

@@

Let us consider two probability spaces: $(\X,\B(\X),\Q)$ and $(\X,\B(\X),\M)$ and let the measurable function $w:\X \rightarrow \R$ be proportional to the Radom-Nikodym derivative between $\Q$ and $\M$, i.e.,  $w(x)\propto\frac{\Q(dx)}{\M(dx)}$. \
The following result represents the starting point for the idea of importance sampling.

\begin{lemma}\label{lemma_imp_sam_1}
For any measurable, integrable and bounded function $\varphi: \X \rightarrow \R$, the following holds:
\begin{equation}
    \E_\M(\varphi \cdot w)=\E_\Q(\varphi)\cdot \E_\M(w)
\end{equation}
\end{lemma}

@@

A second approach can be used when sampling has to be done sequentially: \textit{sequential importance resampling} (SIR). Differently from SIS, an intermediate resampling stage takes place, in which time 0 values are sampled with replacement according to a given resampling strategy (usually, sampling indexes from a multinomial distribution with probabilities equal to the weights computed through IS at time 0).\
While in the SIS all the particles (i.e., the sampled values) ``survive", with the ``risk" that particles associated with small weights are carried-over (in the sense that sampling in period 1 occurs from the kernel $M_1(X_0^n,dx_1)$), the same does not occur in SIR. In this second algorithm, particles with larger weights are more likely to be resampled and be used to ``generate" the particles in the next period.\
We describe the SIR more formally, focusing on the previous two-period case.
\begin{enumerate}
    \item Sample $N$ values $X_0$ from the proposal measure $\M_0$: $X_0^n\thicksim \M_0$ for $n=1,...,N$.
    \item Compute the weights $W_0^n$ for $n=1,...,N$.
    \item Sample $N$ indexes $(A_1^1,...A_1^N)$ from a multinomial distribution such that for all $j=1,...N$, $Pr(\{A_1^j=n\})=W_0^n$, for all $n=1,...N$.
    \item Build a new time 0 sample, using the values for $X_0$ associated to the previously extracted indexes $(A_1^n)_{n}$: $(X_0^{A_1^1},...,X_0^{A_1^N})$.
    \item Sample $N$ values for $X_1$ from the transition kernel $M_1$: $X_1^n\thicksim M_1(X_0^{A_1^n},dx_1)$ for $n=1,...,N$.
\end{enumerate}
Steps 3-4 constitute the actual resampling ones. At this point, let us define the new two-period sequence $\tilde X_{0:1}^n:=(X_0^{A_1^n},X_1^n)$. Then, the particle approximation for the joint distribution of $(X_0,X_1)$ becomes:
\begin{equation*}
     \Q_{1}^N(dx_{0:1})=\frac{1}{N}\sum_{n=1}^N\delta_{\tilde X_{0:1}^n}(dx_{0:1})
\end{equation*}
Note that, after resampling occurred, each particle $\tilde X_{0:1}^n:=(X_0^{A_1^n},X_1^n)$ has the same probability weight.\
To understand the reason of the previous statement, let us assume that both $\Q_0$ and $\M_0$ admit densities $q_0$, $m_0$ with respect to the same dominating measure $\nu(dx_0)$ and that the transition kernel $M_1$ also admits the density $m_1$. 
Consider the sequential updating for the weights:
\begin{equation}
    w^n_1=\frac{\Q_1(\tilde dx_{0:1})}{\M_1(dx_{0:1})}\propto \frac{q_0(x_0^{A_1^n})}{Pr(A_1^n)\cdot m_0(x_0^{A_1^n})}\cdot \frac{m_1(x_1^n|x_0^{A_1^n})}{m_1(x_1^n|x_0^{A_1^n})}
=\underbrace{\frac{q_0(x_0^{A_1^n})}{ m_0(x_0^{A_1^n})}}_{W_0^n}\cdot \underbrace{\frac{1}{Pr(A_1^n)}}_{1/W_0^n}=1\label{sis_update}
\end{equation}
where $\M_1$ denotes the probability measure for the extraction of the particle $\tilde x_{0:1}^n$ according to SIR. $\M_1$ (second equality sign in the previous formula) can be rewritten as the product between the kernel for $x_1^n$ given $x_0^{A_1^n}$ and the probability of $x_0^{A_1^n}$ itself. This last probability can also be rewritten as the product between the proposal density $m_0$ from which $X_0$ is extracted and the probability that it is sampled again when resampling occurs.

\subsection{SIS vs SIR}
The choice between the two sequential sampling strategies (SIS and SIR) has significant implications in terms of computational cost and variance of the importance weights. In fact, it is clear that SIR is more costly computationally since it requires an intermediate resampling step. Nonetheless, SIS suffers from the problem knonw as \textit{curse of dimensionality}, i.e., the fact that the $ESS$ collapses over time as the variance of the weights diverges.\
Consider an extension of SIS to a $T$-period model and let us assume that the target and the proposal kernels admit densities, respectively, $q_t$ and $m_t$ with respect to the same dominating measures (this implies that the joint distributions\footnote{Let us denote the joint distributions with $q_t$ and $m_t$ too, with some abuse of notation} admit a density too). Then, weights, for every $t=1,...,T$, are updated according to:
\begin{equation}
    w_t^n\propto \frac{q_t(x_{0:t}^n)}{m_t(x_{0:t}^n)}\propto \underbrace{\frac{q_{t-1}( x_{0:t-1}^n)}{m_{t-1}(x_{0:t-1}^n)}}_{\propto w_{t-1}^n}\cdot \frac{q_t( x_t|x_{t-1}^n)}{m_t(x_t|x_{t-1}^n)}\propto 
    w_{t-1}^n\cdot\frac{q_t( x_t|x_{t-1}^n)}{m_t(x_t|x_{t-1}^n)}\label{sis_update}
\end{equation}
Equation \eqref{sis_update} suggests, intuitively, that over time new variance is added from the incremental weights (i.e., the updating factor $q_t( x_t|x_{t-1}^n)/m_t( x_t|x_{t-1}^n)$) determining the curse of dimensionality issue. Instead, when resampling takes place, since the past weights $w_{t-1}$ are reset to be equal across them all the variance of $w_t$ inherited from the past is shut-down and the only source of variability at each stage comes from the incremental weights.

@@

\color{red}
Let us consider the standard state-space model introduced in the previous chapters.
\color{black}
\textit{Particle filter} (PF) is a numerical algorithm that is used to obtain approximations for the filtering - $\Pd_t (dx_t|y_{1:t})$ - and the prediction - $\Pd_t (dx_t|y_{1:t-1})$ - distributions for inference in state-space models.\ \
In a state-space model sampling sequentially from the transition kernels for the states is not enough to obtain good approximations since the observation sample ($y_{1:T}$) contains information about the underlying states through the likelihood function $f_t(y_t|x_t)$. Observations constitute signals for the model's states.\
Therefore, the conditional distribution of the states, given the observed data - $\Pd_t(dx_{1:t}|y_{1:t})$ - changes over time as new data points become available. The PF algorithm allows to incorporate sequentially the new data to obtain new approximations for this distribution (and for the filtering and prediction ones). \
In general, at each step of the algorithm, states are sampled sequentially from the proposal measure, call it $\M_t$, and importance auto-normalized weights are computed as the Radom-Nikodym derivative of the target $\Pd_t$ with respect to the proposal $\M_t$. Moreover, the Markov properties of the sequence of the states allows to update sequentially the importance weights.\ \
We will analyse three different types of particle filtering procedures: \textit{bootstrap}, \textit{guided} and \textit{auxiliary particle filters}, that differ between them with respect to their scope and efficiency.

@@

\subsection{Implementation}
Consider again the random walk plus noise model of section \ref{implementation},the challenge here is to estimate the filtered states using a Sequential Importance Sampling algorithm. The main idea of the SIS applied to this univariate linear gaussian model is described in the following algorithm. We indicate with $n$ the sample size of time observations and with $N$ the generated sample size for each step of the Sequenial Monte Carlo. 
\begin{algorithm} SIS filter for Random Walk plus Noise Model
\begin{itemize}
\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
\item For $t=1,...,n$:
\begin{enumerate}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
\end{enumerate}
\end{itemize}
\end{algorithm}
Our \code{SISfun} function implemented in R replicate this steps.
\
\hrule
\hrule
\
\code{SISfun(data,N,m0,C0,tau,sigma)}
\
\hrule
\textbf{Arguments}

@@

It is not always possible to sample sequentially from the correct transition kernel. Moreover, even if it were possible, there exist a proposal transition kernel that makes the algorithm more efficient. This motivates the \textit{guided particle filter} (GPF), that can be interpreted as a more general version of the GPF. In fact, with the GPF it is possible, at each stage, to sample new states from a transition kernel that differs from the correct one.\ \

>>>>

\section{State-Space Models}\label{sec_state_space}
We now define \textit{State-Space models} as Markov processes that are partially observed. Let $X_t\in\X$ and $Y_t\in\Y$ and let $\{(X_t,Y_t)\}$ be a stochastic process. To help intuition, suppose that $\{Y_t\}$ is observable, i.e., coincides with a sequence of observations, while $\{X_t\}$ is latent. Consider the initial distribution $\P_0(dx_0)$ and the probability kernels
$P_t(x_{t-1},dx_t)$ and $F_t(x_t,d_t)$, where $t=1,2,\dots$.
\
\begin{definition}\label{state-space-def}
The process $\{(X_t,Y_t)\}$ is a State-Space model if the joint distribution of $(X_{0:T},Y_{0:T})$ is 
\begin{align*}
  \P_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T}) &= \P_0(dx_0)\prod\limits_{t=1}^T P_t(x_{t-1},dx_t)\prod\limits_{t=1}^T F_t(x_t,dy_t)\\
  &= [\P_0(dx_0)F_0(x_0,y_0)]\prod\limits_{t=1}^T[P_t(x_{t-1},dx_t)F_t(x_t,dy_t)] \\
  &= \P_T(dx_{0:T})\prod \limits_{t=1}^T F_t(x_t,dy_t).
\end{align*}
\end{definition}
From the second inequality, it is apparent that $\{(X_t,Y_t)\}$ is a Markov process with initial distribution $[\P_0(dx_0)F_0(x_0,y_0)]$ and transition kernel $P_t(x_{t-1},dx_t)F_t(x_t,dy_t).$ The third equality, instead, shows that, $\{X_t\}$ has a marginal distribution $$\P_T(dx_{0:T})=\P_0(dx_0)\prod_{t=1}^T P_t(x_{t-1},dx_t),$$ meaning that $\{X_t\}$ is marginally distributed as a Markov process with initial distribution $\P_T(dx_{0:T})$ and transition kernel $P_t(x_{t-1},dx_t).$

Assume now that there exists a dominating measures $\nu(dy)$ such that 
$F_t(x_t,dy_t)=f_t(y_t\mid x_t)\nu(dy_t)$ for all $t.$
Then by the last equality of definition \ref{state-space-def}, the joint distribution of $(X_{0:T},Y_{0:T})$ is 
\begin{equation}\label{state-space-densities}
   \P_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T})=\P_T(dx_{0:T})\prod\limits_{t=0}^T f_t(y_t\mid x_t)\prod\limits_{t=0}^T\nu(dy_t).
\end{equation}
Integrating away $X_{0:T},$ one gets the marginal density
\begin{align*}\label{state-space-marginal-y}
  \P_T(dy_{0:T}) &= \E_{\P_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\nu(dy_s) \\
                &= p_t(y_{0:t})\prod\limits_{s=0}^t\nu(dy_s),
\end{align*}                
where $p_t(y_{0:t})$ is the likelihood function.
Dividing the left-hand side of equation \ref{state-space-densities} by the left-hand side of equation \ref{state-space-marginal-y} and the right-hand side of equation \ref{state-space-densities} by the right-hand side of equation \ref{state-space-marginal-y}, one obtains the filtering distribution
\begin{equation}
\P_t(X_{t}\in dx_{t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\P_t(dx_{0:t}).
\end{equation}

@@

Suppose that $\X=\R^{d_x}$ and $\Y=\R^{d_y}.$ Let $U_t\thicksim\N(0,\Sigma_t)$ and $V_t\thicksim\N(0,R_t)$ be two independent white-noise errors and assume the initial distribution 
$X_0\thicksim\N(0,\Sigma_0).$ We have a linear Gaussian model if 
\begin{align}
X_t &= A_tX_{t-1}+U_t \\
Y_t &= B_tX_t + V_t.
\end{align}
Note that  $X_t\mid X_{t-1}$ and $Y_t\mid X_t$ are both normally distributed. 
\
\begin{lemma}\label{kalman-filter}
  Let $X\thicksim\N(m_0,Q_0)$ and $Y\mid X\thicksim\N(BX,R)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
  Then, it holds that
  \begin{align}
    X\mid Y=y&\thicksim\N(m_1,Q_1),\\
    Q_1&= Q_0 (I_{d_x} - B'(B Q_0 B'+ R)^{-1} B Q_0) \\
    m_1&=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
  \end{align}
\end{lemma}
\begin{proof*}
   The lemma can be proved using Bayes' Theorem. Suppose $X\thicksim\N(m_0, Q_0).$ It is easy to see that
\begin{equation} \label{normal-distrib-rewrite}
        p(x)\propto \text{exp}\Big(-\frac{1}{2}X'AX + X'B\Big),
\end{equation}
with $A=Q_0^{-1}$ and $B=Am_0.$
Now, by Bayes' Theorem,

    \begin{align*}
        p(x\mid Y=y)  &\propto p(x) p(y\mid x) \\
                    &\propto \text{exp}\Big(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}(y-BX)'R^{-1}(y-BX)) \\
                    &\propto \text{exp}(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}X' B' R^{-1}BX+X' B' R^{-1}y)\\
                    &=\text{exp}(-\frac{1}{2}X'(Q_0^{-1}+B' R^{-1}B)X+
                    X'(Q_0^{-1}m_0+B' R^{-1}y)\Big).
    \end{align*}

Hence, it follows that 
\begin{equation*}
    Q_1=Var(X\mid Y=y)= (Q_0^{-1}+ B' R^{-1} B)^{-1}=Q_0-Q_0 B'(B Q_0 B'+R)^{-1}B Q_0
\end{equation*}
and 
    \begin{align*}
        m_1 &=\E(X\mid Y=y) = Q_1^{-1}(Q_0^{-1}m_0+B' R^{-1}y)\\
            &= (Q_0-Q_0B'(B Q_0B'+R)^{-1}B Q_0)(Q_0^{-1}m_0+B' R^{-1}y)\\
            &= m_0-Q_0B'(B Q_0B'+R)^{-1}B m_0 + Q_0 B' R^{-1}y - Q_0B'(B Q_0B'+R)^{-1}B Q_0 B' R^{-1}y \\
            &=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
    \end{align*}
\end{proof*}

Assume that $\P_{t-1}(dx_{t-1}\mid Y_{0:t-1}=y_{0:t-1})=\N(m_{t-1},Q_{t-1}).$
Then, using the transition equation of the latent variable, 
by the linearity of the Gaussian distribution, we have that the predictive distribution is 
\begin{equation}
\P_{t-1}(dx_{t}\mid Y_{0:t-1}=y_{0:t-1})=\N(A_tm_{t-1},A_tQ_{t-1}A_t'+\Sigma_t).
\end{equation}
Letting $E_T=A_tQ_{t-1}A_t'+\Sigma_t,$ and applying lemma \ref{kalman-filter}, it follows that the filtering distribution is 
\begin{align*}
  \P_{t}(X_t\in dx_{t}\mid Y_{0:t-1}&=y_{0:t-1})=\N(m_t,Q_t), \text{where} \\ Q_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
  m_t &= [I_{d_x} - E_tB_t'(B_tE_tB_t'+R_t)^{-1} B_t]A_tm_{t-1}+E_tB_t'(B_tE_tB_t'+R_t)^{-1}y_t.
\end{align*}

@@

Let us consider two probability spaces: $(\X,\B(\X),\Q)$ and $(\X,\B(\X),\M)$ and let the measurable function $w:\X \rightarrow \R$ be proportional to the Radom-Nikodym derivative between $\Q$ and $\M$, i.e.,  $w(x)\propto\frac{\Q(dx)}{\M(dx)}$. \
The following result represents the starting point for the idea of importance sampling.
\bigskip
\begin{lemma}\label{lemma_imp_sam_1}
For any measurable, integrable and bounded function $\varphi: \X \rightarrow \R$, the following holds:
\begin{equation}
    \E_\M(\varphi \cdot w)=\E_\Q(\varphi)\cdot \E_\M(w)
\end{equation}
\end{lemma}

@@

A second approach can be used when sampling has to be done sequentially: \textit{sequential importance resampling} (SIR). Differently from SIS, an intermediate resampling stage takes place, in which time 0 values are sampled with replacement according to a given resampling strategy (usually, sampling indexes from a multinomial distribution with probabilities equal to the weights computed through IS at time 0).\
While in the SIS all the particles (i.e., the sampled values) "survive", with the "risk" that particles associated with small weights are carried-over (in the sense that sampling in period 1 occurs from the kernel $M_1(X_0^n,dx_1)$), the same does not occur in SIR. In this second algorithm, particles with larger weights are more likely to be resampled and be used to ``generate" the particles in the next period.\
We describe the SIR more formally, focusing on the previous two-period case.
\begin{enumerate}
    \item Sample $N$ values $X_0$ from the proposal measure $\M_0$: $X_0^n\thicksim \M_0$ for $n=1,...,N$.
    \item Compute the weights $W_0^n$ for $n=1,...,N$.
    \item Sample $N$ indexes $(A_1^1,...A_1^N)$ from a multinomial distribution such that for all $j=1,...N$, $Pr(\{A_1^j=n\})=W_0^n$, for all $n=1,...N$.
    \item Build a new time 0 sample, using the values for $X_0$ associated to the previously extracted indexes $(A_1^n)_{n}$: $(X_0^{A_1^1},...,X_0^{A_1^N})$.
    \item Sample $N$ values for $X_1$ from the transition kernel $M_1$: $X_1^n\thicksim M_1(X_0^{A_1^n},dx_1)$ for $n=1,...,N$.
\end{enumerate}
Steps 3-4 constitute the actual resampling ones. At this point, let us define the new two-period sequence $\tilde X_{0:1}^n:=(X_0^{A_1^n},X_1^n)$. Then, the particle approximation for the joint distribution of $(X_0,X_1)$ becomes:
\begin{equation*}
     \Q_{1}^N(dx_{0:1})=\frac{1}{N}\sum_{n=1}^N\delta_{\tilde X_{0:1}^n}(dx_{0:1})
\end{equation*}
Note that, after resampling occurred, each particle $\tilde X_{0:1}^n:=(X_0^{A_1^n},X_1^n)$ has the same probability weight.\
To understand the reason of the previous statement, let us assume that both $\Q_0$ and $\M_0$ admit densities $q_0$, $m_0$ with respect to the same dominating measure $\nu(dx_0)$ and that the transition kernel $M_1$ also admits the density $m_1$. 
Consider the sequential updating for the weights:
\begin{equation}
    w^n_1=\frac{\Q_1(dx_{0:1})}{\M_1(dx_{0:1})}\propto \frac{q_0(x_0^{A_1^n})}{Pr(A_1^n)\cdot m_0(x_0^{A_1^n})}\cdot \frac{m_1(x_1^n|x_0^{A_1^n})}{m_1(x_1^n|x_0^{A_1^n})}
=\underbrace{\frac{q_0(x_0^{A_1^n})}{ m_0(x_0^{A_1^n})}}_{W_0^n}\cdot \underbrace{\frac{1}{Pr(A_1^n)}}_{1/W_0^n}=1\label{sis_update}
\end{equation}
where $\M_1$ denotes the probability measure for the extraction of the particle $\tilde x_{0:1}^n$ according to SIR. $\M_1$ (second equality sign in the previous formula) can be rewritten as the product between the kernel for $x_1^n$ given $x_0^{A_1^n}$ and the probability of $x_0^{A_1^n}$ itself. This last probability can also be rewritten as the product between the proposal density $m_0$ from which $X_0$ is extracted and the probability that it is sampled again when resampling occurs.

\subsection{SIS vs SIR}
The choice between the two sequential sampling strategies (SIS and SIR) has significant implications in terms of computational cost and variance of the importance weights. In fact, it is clear that SIR is more costly computationally since it requires an intermediate resampling step. Nonetheless, SIS suffers from the problem knonw as \textit{curse of dimensionality}, i.e., the fact that the $ESS$ collapses over time as the variance of the weights diverges.\
Consider an extension of SIS to a $T$-period model and let us assume that the target and the proposal kernels admit densities, respectively, $q_t$ and $m_t$ with respect to the same dominating measures (this implies that the joint distributions\footnote{Let us denote the joint distributions with $q_t$ and $m_t$ too, with some abuse of notation.} admit a density too). Then, weights, for every $t=1,...,T$, are updated according to:
\begin{equation}
    w_t^n\propto \frac{q_t(x_{0:t}^n)}{m_t(x_{0:t}^n)}\propto \underbrace{\frac{q_{t-1}( x_{0:t-1}^n)}{m_{t-1}(x_{0:t-1}^n)}}_{\propto w_{t-1}^n}\cdot \frac{q_t( x_t|x_{t-1}^n)}{m_t(x_t|x_{t-1}^n)}\propto 
    w_{t-1}^n\cdot\frac{q_t( x_t|x_{t-1}^n)}{m_t(x_t|x_{t-1}^n)}\label{sis_update}
\end{equation}
Equation \eqref{sis_update} suggests, intuitively, that over time new variance is added from the incremental weights (i.e., the updating factor $q_t( x_t|x_{t-1}^n)/m_t( x_t|x_{t-1}^n)$) determining the curse of dimensionality issue. Instead, when resampling takes place, since the past weights $w_{t-1}$ are reset to be equal across them all the variance of $w_t$ inherited from the past is shut-down and the only source of variability at each stage comes from the incremental weights.

@@

Let us consider the standard state-space model introduced previously (see \ref{sec_state_space}).
\textit{Particle filter} (PF) is a numerical algorithm that is used to obtain approximations for the filtering - $\Pd_t (dx_t|y_{1:t})$ - and the prediction - $\Pd_t (dx_t|y_{1:t-1})$ - distributions for inference in state-space models.\ \
In a state-space model sampling sequentially from the transition kernels for the states is not enough to obtain good approximations since the observation sample ($y_{1:T}$) contains information about the underlying states through the likelihood function $f_t(y_t|x_t)$. Observations constitute signals for the model's states.\
Therefore, the conditional distribution of the states, given the observed data - $\Pd_t(dx_{1:t}|y_{1:t})$ - changes over time as new data points become available. The PF algorithm allows to incorporate sequentially the new data to obtain new approximations for this distribution (and for the filtering and prediction ones). \
In general, at each step of the algorithm, states are sampled sequentially from the proposal measure, call it $\M_t$, and importance auto-normalized weights are computed as the Radom-Nikodym derivative of the target $\Pd_t$ with respect to the proposal $\M_t$. Moreover, the Markov properties of the sequence of the states allows to update sequentially the importance weights.\ \
We will analyse three different types of particle filtering procedures: \textit{bootstrap}, \textit{guided} and \textit{auxiliary particle filters}, that differ between them with respect to their scope and efficiency.

@@

\subsection{Implementation}
Consider again the random walk plus noise model of section \ref{implementation},the challenge here is to estimate the filtered states using a Sequential Importance Sampling algorithm. The main idea of the SIS applied to this univariate linear gaussian model is described in the following algorithm. We indicate with $n$ the sample size of time observations and with $N$ the generated sample size for each step of the Sequenial Monte Carlo. 
\begin{algorithm} SIS filter for Random Walk plus Noise Model
\begin{itemize}
\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarize $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
\item For $t=1,...,n$:
\begin{enumerate}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
\end{enumerate}
\end{itemize}
\end{algorithm}
Our \code{SISfun} function implemented in R replicate this steps.
\
\hrule
\hrule
\
\code{SISfun(data,N,m0,C0,tau,sigma)}
\
\hrule
\textbf{Arguments}

@@

It is not always possible to sample sequentially from the correct transition kernel. Moreover, even if it were possible, there exist a proposal transition kernel that makes the algorithm more efficient. This motivates the \textit{guided particle filter} (GPF), that can be interpreted as a more general version of the BPF. In fact, with the GPF it is possible, at each stage, to sample new states from a transition kernel that differs from the correct one.\ \
